When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-01-19 doc/paper/vpd_et_paper_PG_v1.tex
good progress

\subsection{Functional Form of the Sign Term}
\label{sign_func}
First, we explore the variables within the sign term to gain better intuition on the driver of either the increase or reduction of ET with VPD. $c_s$ and $\gamma$ are relatively constant to a good approximation so that the variability is dominated by $\sigma$ and $VPD$. $uWUE$ could vary with soil moisture but has been shown to be relatively constant except in very dry soil conditions \citep{Zhou_2016, Lin_2015}. We thus further assume that the soil is not too dry so that $uWUE$ can be approximated as a constant. This then means that the sign term only depends on VPD for a given PFT and is approximately just a function of $VPD$. We can further determine a critical threshold operating increase from decrease in ET, such that $VPD_{crit}$ where $\frac{\partial \; ET}{\partial \; VPD} = 0$:

@@

Based on Figure \ref{term3} and Table \ref{vpd_crit}, CROs are the least water conservative and have the most positive constant offset, while CSH are the most water conservative and have the most negative constant offset in the sign term.  ENF ($g_1 = 74.31$) has by far the largest VPD dependence of response, while CRO ($g_1 = 183.1$) has the smallest VPD dependence. This means ENF's stomatal response evolved to change substantially depending on the envionmental dryness. This contasts\explain{think about this - what does it mean - go back to Franks and think about g1's relation to wue}

A primary takeaway from Figure \ref{term3}a is that according to our theory for all PFTs except for crops there is frequent occurrence of a negative (plant dominating) ET response to increases in VPD. This thus means that plants are able in most atmospheric conditions to redue ET in response to increased VPD and thus to reduce water losses. To better illustrate this, the ranges of observed environmental VPDs at the FLUXNET sites are plotted parallel to the x-axis. ALL OF THIS SHOULD BE IN CAPTION NOT IN MAIN TEXT: (25th, 50th, and 75th percentiles are marked by stars), 
The location of VPD$_{crit}$ is marked by a vertical line. For CSH, VPD is always less than VPD$_{crit}$ so that the plant response dominates and empathizes the water conservative strategy of those plants. For CRO on the other, VPD is almost always higher than VPD$_{crit}$, emphasizing that those plants are water intensive and were actually engineered for photosynthesis rather than water saving. For forests and grass, whether plant response or atmospheric demand dominates is more dependent on the environment, with about half of of observed VPD less than VPD$_{crit}$, i.e. in conditions where plant response dominates. It is also important to note that even when atmospheric demand dominates, the response is still far smaller than it would be for potential evaporation i.e. atmospheric demand only, emphasizing that there is still a strong regulation of evaporative flux by stomata and though the plant xylem. The sign term in this case would just be a constant ($\frac{c_p}{R_{air}} \approx 3.5$), which is far larger than any part of the curves for any PFT. This highlights the deficiencies of PET's ability to capture land response to changes in atmospheric dryness (VPD) SEEMS TOO GENERIC BETTER EXPLAIN. However, we have assumed $\sigma$ (uncertainty) equal to one. We still need to test if inclusion of uncertainty could change our conclusions.
\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig05.pdf}
\caption{Sources of variability for the sign term. Top: sign term as a function of VPD, with $\sigma$ held fixed at 1. The observed range of VPD for each PFT is also shown below the x-axis. Line extent corresponds to 5th and 95th percentiles, while stars denote the location of the 25th, 50th, and 75th percentiles.

@@

Figure \ref{term3}b shows the location of VPD$_{crit}$, as a function of $\sigma$ (uncertainty) and $VPD$. For any $\sigma$ or VPD less (more) than these curves, the sign term will be negative (positive). It is clear that the portion of VPD observations below/above these curves will be a strong function of $\sigma$. However, we can see some general trends. For CSH, $\frac{\partial \; ET}{\partial \; VPD}$ should be negative for the vast majority of observed $\sigma$ and VPD. The fraction of positive $\frac{\partial \; ET}{\partial \; VPD}$ appears to be more even for ENF, GRA, and DBF, and we might expect a greater frequency of positive $\frac{\partial \; ET}{\partial \; VPD}$ for CRO. Section \ref{testing} further examines the consistency of our theory, given  the apparent sensitivity to uncertainty ($\sigma$). THIS PARAGRAPH DOES NOT SEEM CONCLUSIVE AND MORE A TRAIN OF THOUGHT. LAY DOWN EXACTLY WHAT YOU HAVE IN MIND.

@@

While the above discussion of the sign of $\frac{\partial \; ET}{\partial \; VPD}$ is important to answer our research question, the magnitude of $\frac{\partial \; ET}{\partial \; VPD}$ will also the relative magnitude of the change $\frac{\partial \; ET}{\partial \; VPD}$ and the importance of $VPD$ variability for overall $ET$ variability. So we now more closely examine the scaling term: $\frac{P}{T} \propto \rho$. Air density $\rho$ scaling term varies little relative to aerodynamic conductance and $\Delta$. The psychrometric constant $\gamma$ is also relatively constant, so the scaling term should be primarily a function of aerodynamic conductance and temperature, through the slope of the Clauisu-Clapeyron relationship $\Delta$. This is as expected, given that the aerodynamic conductance represents the efficiency fo the transfer of surface anomalies to the atmosphere. As it increases I THINK YOU REALLY MEAN CONDUCATANCE HERE - BE CAREFUL, any plant response will be communicated more strongly to the atmosphere (and vice-versa).

REMOVE ANY SNETENCES LIKE THIS MAKES SENSES ETC...

@@

Theoretical derivation has so far illuminated the dependence of $\frac{\partial \; ET}{\partial \; VPD}$ and how it varies across PFT. Large mean $uWUE$ shifts CRO and DBF towards positive $\frac{\partial \; ET}{\partial \; VPD}$ i.e. increasing ET at higher VPD, but $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ is still negative for DBF due to its lower $g_1$ SAY PHISICALLY WHAT IT IS FOR THE READER TO UNDERSTAND. ENF's low $g_1$, which is even lower than DBF, increases the dependence of $\frac{\partial \; ET}{\partial \; VPD}$ on $VPD$, and makes the function strongly nonlinear. This has the side effect of pushing $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ negative further than other PFTs for a given fraction $\frac{\partial \; ET}{\partial \; VPD} < 0$ and magnitude $\frac{\partial \; ET}{\partial \; VPD}(\overline{T,\ldots,VPD})$. SAY THIS IN WORDS AS WELL OR YOU WILL LSOE THE READER. GRA shows the opposite behavior; a relatively high $g_1$  makes the function more linear, decreasing the magnitude of $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ for a given fraction $\frac{\partial \; ET}{\partial \; VPD} < 0$ and negative $\frac{\partial \; ET}{\partial \; VPD}(\overline{T,\ldots,VPD})$. SAY IN WORDS AGAIN WHAT IT MEANS OTHERWISE PEOPLE WILL NOT UNDERSTAND OUR RESULTS. Finally, the low $uWUE$ of CSH pushes it toward the lowest values of $\frac{\partial \; ET}{\partial \; VPD}$ (Figure \ref{term3}) across all PFTs WHAT DOES IT MEAN AGAIN. Variability in $VPD$ accounts for the largest amount of $ET$ variability for CSH. For the other PFTs, $VPD$ contributes less to $ET$ variability, but still represents about 30-40 \% of $R$'s contributions to ET variability. WHAT IS THE SORUCE OF VARIBILITY OF THE REST OF ET I DON"T KNOW IF THAT IS RELEVANT FOR THE PRESENT PAPER.

@@

\subsection{Validation of theory at eddy-covariance sites}
\label{testing}
So far we have developed a theory for $\frac{\partial \; ET}{\partial \; VPD}$'s behavior as a function of VPD and PFTs. We now turn to an direct evaluation against observations taken from a wide range of eddy-covariance datasets (SEE METHODS).

>>>>

\subsection{Functional Form of the Sign Term}
\label{sign_func}
First, we explore the variables within the sign term to gain better intuition on the driver of either the increase or reduction of ET with VPD. $c_s$ and $\gamma$ are relatively constant to a good approximation so that the variability is dominated by $\sigma$ and $VPD$. $uWUE$ could vary with soil moisture but has been shown to be relatively constant except in very dry soil conditions \citep{Zhou_2016, Lin_2015}. We thus further assume that the soil is not too dry so that $uWUE$ can be approximated as a constant. This then means that the sign term only depends on VPD for a given PFT and is approximately just a function of $VPD$. We can further determine a critical threshold separating an increase from decrease in ET, such that $VPD_{crit}$ where $\frac{\partial \; ET}{\partial \; VPD} = 0$:

@@

Based on Figure \ref{term3} and Table \ref{vpd_crit}, CROs are the least water conservative and have the most positive constant offset, while CSH are the most water conservative and have the most negative constant offset in the sign term.  ENF ($g_1 = 74.31$) has by far the largest VPD dependence of response, while CRO ($g_1 = 183.1$) has the smallest VPD dependence. ENF are less willing to trade water for new growth, so the regulate water much more strongly depending on the dryness environment (VPD). CRO on the otherhand are more willing to trade water for new growth, so they regulate water relatively similiarly irregardless of the environment.

A primary takeaway from Figure \ref{term3}a is that according to our theory for all PFTs except for crops there is frequent occurrence of a negative (plant dominating) ET response to increases in VPD. This thus means that plants are able in most atmospheric conditions to redue ET in response to increased VPD and thus to reduce water losses. To better illustrate this, the ranges of observed environmental VPDs at the FLUXNET sites are plotted parallel to the x-axis. For CSH, VPD is always less than VPD$_{crit}$ so that the plant response dominates and empathizes the water conservative strategy of those plants. For CRO on the other, VPD is almost always higher than VPD$_{crit}$, emphasizing that those plants are water intensive and were actually engineered for photosynthesis rather than water saving. For forests and grass, whether plant response or atmospheric demand dominates is more dependent on the environment, with about half of of observed VPD less than VPD$_{crit}$, i.e. in conditions where plant response dominates. It is also important to note that even when atmospheric demand dominates, the response is still far smaller than it would be for potential evaporation i.e. atmospheric demand only, emphasizing that there is still a strong regulation of evaporative flux by stomata and though the plant xylem. The sign term in this case would just be a constant ($\frac{c_p}{R_{air}} \approx 3.5$), which is far larger than any part of the curves for any PFT. This highlights the deficiencies of PET's ability to capture land response to changes in atmospheric dryness (VPD). Plants are always regulating water exchange from the land surface, even when they reach the limits of they ability to do so. Therefore, the actual land surface reponse to a change in VPD will always me more negative than if the plants are not included in the physics (PET). However, we have assumed $\sigma$ (uncertainty) equal to one. We still need to test if inclusion of uncertainty could change our conclusions.
\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig05.pdf}
\caption{Sources of variability for the sign term. Top: sign term as a function of VPD, with $\sigma$ held fixed at 1. The observed range of VPD for each PFT is also shown below the x-axis. Line extent corresponds to 5th and 95th percentiles, while stars denote the location of the 25th, 50th, and 75th percentiles.

@@

Figure \ref{term3}b shows the location of VPD$_{crit}$, as a function of $\sigma$ (uncertainty) and $VPD$. For any $\sigma$ or VPD less (more) than these curves, the sign term will be negative (positive). It is clear that the portion of VPD observations below/above these curves will be a strong function of $\sigma$. This motivates a thorough examination of how inclusion of uncertainty ($\sigma$) might change the form of the sign term, and in particular the location of VPD$_{crit)$ for each of our plant types. In Secion \ref{testing} we examine if and how uncertainty changes the analysis presented in this section, for each PFT.

@@

While the above discussion of the sign of $\frac{\partial \; ET}{\partial \; VPD}$ is important to answer our research question, the magnitude of $\frac{\partial \; ET}{\partial \; VPD}$ will also the relative magnitude of the change $\frac{\partial \; ET}{\partial \; VPD}$ and the importance of $VPD$ variability for overall $ET$ variability. So we now more closely examine the scaling term: $\frac{P}{T} \propto \rho$. Air density $\rho$ scaling term varies little relative to aerodynamic conductance and $\Delta$. The psychrometric constant $\gamma$ is also relatively constant, so the scaling term should be primarily a function of aerodynamic conductance and temperature, through the slope of the Clauisu-Clapeyron relationship $\Delta$. This is as expected, given that the aerodynamic conductance represents the efficiency fo the transfer of surface anomalies to the atmosphere. As aerodynamic conductace  increases, any plant response will be communicated more strongly to the atmosphere (and vice-versa).

@@

Theoretical derivation has so far illuminated the dependence of $\frac{\partial \; ET}{\partial \; VPD}$ and how it varies across PFT. Large mean $uWUE$ shifts CRO and DBF towards a tendancy to increase ET with VPD (positive $\frac{\partial \; ET}{\partial \; VPD}$), but a lower slope parameter ($g_1$) for DBF allows for decreaes in ET with VPD, at low environmental VPD. \explain{physically state how slope parameter is trade for water at leaf level, but uWUE is trade-ff between GPP and VPD at ecosystem-level}.  ENF's very low slope parameter ($g_1$), which is even lower than for DBF, increases the dependence of ET response  ($\frac{\partial \; ET}{\partial \; VPD}$) on $VPD$, and makes the function strongly nonlinear. This has the effect of making mean ET response ($\overline{\frac{\partial \; ET}{\partial \; VPD}}$)  more negative for a given frequency of occurence of negative ET response ($\frac{\partial \; ET}{\partial \; VPD} < 0$), than for other PFT. GRA shows the opposite behavior; a relatively high slope parameter ($g_1$)  makes the function more linear, causing a more positive mean ET reponse ($\overline{\frac{\partial \; ET}{\partial \; VPD}}$) for a given given frequence of negative ET reponse  ($\frac{\partial \; ET}{\partial \; VPD} < 0$). Finally, CSH's very low $uWUE$ dominates the slope parameter's effects, causing ET to almost always decrees with increasing atmospheric dryness (VPD), even with uncertainty included. Variability in $VPD$ accounts for the largest amount of $ET$ variability for CSH. For the other PFTs, $VPD$ contributes less to $ET$ variability, but still represents about 30-40 \% of $R$'s contributions to ET variability. \explain{think pierre is correct, could probably scrap this anecdote: WHAT IS THE SORUCE OF VARIBILITY OF THE REST OF ET I DON"T KNOW IF THAT IS RELEVANT FOR THE PRESENT PAPER.}

@@

\subsection{Validation of theory at eddy-covariance sites}
\label{testing}
So far we have developed a theory for $\frac{\partial \; ET}{\partial \; VPD}$'s behavior as a function of VPD and PFTs. We now turn to an direct evaluation against observations taken from a wide range of eddy-covariance datasets (Section \ref{methods}).
